Individual and Differential Harm in Redistricting

### 2022-08-30 paper/harm.Rmd
add gill cite
edit

Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
Many factors influence the performance and qualities of an electoral system.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it was impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
Once a set
of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

@@

After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
These represent 44% of all Republicans and 2.5% of all Democrats.
We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than the average Democrat is to be represented by a Republican, compared to plan (b).
In direct terms, this means, on account of their partisan affiliation, Democrats would be 41.5% less likely to be represented by a Democrat than Republicans are to be represented by a Republican.

@@

A voter is harmed only if they are not represented by the candidate of their choice under $q$ (i.e. $\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ (i.e. $\ind\{v(i) = \win_i(v, \tilde q)\}$).

@@

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans which are consistent with districting criteria.
We could also consider averaging over a set of individuals with some shared characteristics.

@@

Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$. Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
In the algorithmic fairness literature, the implicit counterfactual is a perfect predictor and the actual decision process harms individuals whenever it makes a false positive or false negative error.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
To quantify deviations from *fairness*, we introduce the concept of *differential harm*, defined as $$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
$$ for two groups $G_k$ and $G_l$.
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.

@@

For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans which are geographically contiguous, have district populations within a specified deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan or other considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined.
There is certainly an element of subjectivity in sampling a set of districting plans.
Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].

@@

More complicated election outcome models are certainly possible and are straightforward to use with the measures proposed here.
For example, one could operate on the logit scale, model incumbency effects, or add terms to account for local factors.

@@

With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
For each sampled counterfactual districting plan, $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
When other groups (such as racial groups) are used in $\Delta H(q)$, we must repeat these computations for each group.
Finally, we average across all of the samples, and add or subtract across groups, to produce estimates of $E[H(q)]$ and $\Delta_{\dem-\rep} H(q)$.

@@

A primary advantage of our proposed harm measures is that they our measures are defined at the individual level. 
As such, we can study redistricting for any defined group memberships.
For social science, perhaps the two most important group memberships are that of race and party.
For these important contexts, we begin by showing that our measures provide an improvement over existing analytical approaches.

@@

In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
We can account for geographic sorting, political cohesion, and the individual preferences of every voter.
Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.

@@

To apply our proposed framework to study the council district boundaries, we predict individual voting preferences under counterfactual districting plans.
This is more difficult than in both New Jersey and Alabama, for two reasons.
First, the city council is officially nonpartisan, but city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.

@@

Like all ecological inference models, ours relies on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
Yet, the model is a useful simplification of reality which captures the major electoral dynamics relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
The latent factor ecological inference model is not necessary for this kind of analysis; local survey data with ideological preferences and racial information would also suffice.

@@

Our hypothetical elections are focused on cases with two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.

@@

Recall that harm is asymmetric.
While that may be an undesirable property in some areas, here we can leverage that fact.
We flip the setup and compute harm by averaging over the simulated plans with the enacted plan as the counterfactual.
Here, comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
The reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.

@@

The proposed framework is far from the end of the road.
Its flexibility comes at a cost---most notably, the need to specify a set of counterfactual districting plans, the choice of which has a significant impact on conclusions.
Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do.
And while modern simulation methods are a valuable method for defining counterfactuals, these approaches involve analyst discretion and interpretation of often-murky legal standards.
When subjective choices are not represented consistently, different analyses may arrive at different conclusions about which voters are harmed and at what rate.

>>>>

Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
Many factors influence the performance and qualities of an electoral system.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

@@

After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
These represent 44% of all Republicans and 2.5% of all Democrats.
We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.

@@

A voter is harmed only if they are not represented by the candidate of their choice under $q$ ($\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ ($\ind\{v(i) = \win_i(v, \tilde q)\}$).

@@

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
We could also consider averaging over a set of individuals with some shared characteristics.

@@

Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$. Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
There, the implicit counterfactual is a perfect predictor.
Compared to the perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
$$ for two groups $G_k$ and $G_l$.
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.

@@

For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans that are geographically contiguous, have district populations within a specified deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan or other considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined.
There is certainly an element of subjectivity in sampling a set of districting plans.
Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].

@@

More complicated election outcome models are certainly possible and are straightforward to use with the measures proposed here.
For example, one could operate on the logit scale, model incumbency effects, or include additional data from local elections.

@@

With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
For each sampled counterfactual districting plan $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
When other groups (such as racial groups) are used in $\Delta H(q)$, we must repeat these computations for each group.
Finally, we average across all of the samples, and add or subtract across groups, to produce estimates of $E[H(q)]$ and $\Delta_{\dem-\rep} H(q)$.

@@

A primary advantage of our proposed harm measures is that they our measures are defined at the individual level,
allowing for the study of redistricting for any defined group memberships.
For political science, two of the most important group memberships are perhaps race and party.
Even in these well-studied contexts, our measures provide an improvement over existing analytical approaches.

@@

In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
They account for geographic sorting, political cohesion, and the individual preferences of every voter.
Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.

@@

To apply our proposed framework to study the council district boundaries, must be able to we predict individual voting preferences under counterfactual districting plans.
This is more difficult than in both New Jersey and Alabama, for two reasons.
First, the city council is officially nonpartisan, and city politics are dominated by the Democratic Party, so voting choices are more complex than a simple partisan preference.
To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.

@@

Like all ecological inference models, ours relies on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one choice out of many model which could provide estimated vote choices for individuals.
Local survey data with ideological preferences and racial information would be a sufficient alternative.

@@

Our hypothetical elections involve only two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.

@@

These two harm estimates reflect the harm of the current boundaries compared to a race-neutral baseline.
The more policy-relevant direction might be in evaluating a change from the status quo.
What is the harm of switching to race-neutral districting from the current race-aware boundaries?
We can flip the above setup and compute harm by averaging over the simulated plans with the enacted plan as the counterfactual.
For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.

@@

The proposed framework is far from the end of the road.
Its flexibility comes at a cost---most notably, the need to specify a set of counterfactual districting plans, the choice of which has a significant impact on conclusions.
Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do.
And while modern simulation methods are a valuable starting point for defining counterfactuals, these approaches involve analyst discretion and interpretation of often-murky legal standards.
When subjective choices are not represented consistently, different analyses may arrive at different conclusions about which voters are harmed and at what rate.
